---
title: "Final project"
author: "Murtaza Yaqubi"
format: html
editor: visual
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(readr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(naniar)
```

## General Background Information

Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease. It’s also what most people mean when they use the general term “heart disease”. It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart. This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath. As the leading cause of death in the U.S. and globally, CAD claimed 375,500 lives in the U.S. in 2021 and affects over 18 million American adults.

## Description of data and data source

This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses. Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).

This heart disease dataset is acquired from one o f the multispecialty hospitals in India. This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type). This dataset was obtained from mendeley data where it was first published in April of 2021.

This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>

## Questions/Hypotheses to be addressed

Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?

refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820

## Methods

For this analysis I will be using target which is the outcome of interest and is described as abcense or pressence of heart disease and some of the predictors of interest are age, serum cholesterol, resting BP, and old peak. The goal for this analysis is to be able observe if young patients with normal cholesterol but high BP/ST depression have higher CVD rates than expected.

The analysis will first isolate younger individuals with clinically normal cholesterol levels to examine their risk of cardiovascular disease (CVD) when other factors like high blood pressure or ST depression are present. Patients will be categorized based on these secondary risk factors, and descriptive analysis will compare CVD prevalence among groups, such as those with normal cholesterol but elevated blood pressure versus those with entirely normal profiles. This step aims to reveal potential associations between non-traditional risk factors and CVD in younger populations.

To validate these findings, statistical tests like chi-square analysis will assess significance, while logistic regression models will quantify the independent and combined effects of blood pressure and ST depression on CVD risk, adjusting for confounders like gender and fasting blood sugar. Visualization techniques, including bar plots and boxplots, will further illustrate risk patterns, making the findings more interpretable. By integrating these methods, the study aims to challenge traditional risk-assessment frameworks and support more nuanced screening approaches for younger individuals at risk of CVD.

## Data aquisition

This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.

This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>

## Part 1, Data import and cleaning

```{r}
cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))

head(cardio_df)  # View the column names

colnames(cardio_df)  # View the column names

dim(cardio_df)  # View the dimensions of the data

summary(cardio_df)  # View the summary of the data

glimpse(cardio_df)  # Get a glimpse of the dat

gg_miss_var(cardio_df)  # Ilustrate the number of NA's for each of the variables



# Transform variables into suitable types and label them according to the data dictionary to enhance readability.
cardio_df <- cardio_df %>%
  mutate(
    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
  )

print(cardio_df)
```

## Part 2. Exploratory Data Analysis (EDA):

```{r}
# Calculate summary statistics for selected variables

summary_stats <- cardio_df %>% 
  summarise(
    # Age
    Age_min = min(age, na.rm = TRUE),
    Age_max = max(age, na.rm = TRUE),
    Age_mean = mean(age, na.rm = TRUE),
    Age_sd = sd(age, na.rm = TRUE),
    
    # Resting Blood Pressure
    RestingBP_min = min(restingBP, na.rm = TRUE),
    RestingBP_max = max(restingBP, na.rm = TRUE),
    RestingBP_mean = mean(restingBP, na.rm = TRUE),
    RestingBP_sd = sd(restingBP, na.rm = TRUE),
    
    # Serum Cholesterol
    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
    
    # Maximum Heart Rate
    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
    
    # Oldpeak
    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
    Oldpeak_max = max(oldpeak, na.rm = TRUE),
    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
    Oldpeak_sd = sd(oldpeak, na.rm = TRUE),
  )

# Print the summary statistics
print(summary_stats)


# Histogram and density overlay of age distribution
ggplot(cardio_df, aes(x = age)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
  geom_density(color = "darkblue", linewidth = 1) +
  labs(title = "Age Distribution", x = "Age", y = "Density") +
  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
  theme_bw()



# Histogram and density overlay of serum cholestrol (excluding zeros)
ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
  geom_density(color = "darkgreen", linewidth = 1) +
  labs(title = "Serum Cholesterol Distribution", 
       x = "Serum Cholesterol (mg/dL)", y = "Density") +
  theme_minimal()


# Histogram and density overlay of blood pressure (excluding zeros)
ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Resting Blood Pressure Distribution", 
       x = "Resting Blood Pressure (mmHG)", y = "Density") +
  theme_bw()

```

## Part 3, Statistical analysis:

```{r}

# Filter data for younger individuals (age < 40) with normal cholesterol levels, elevated BP and elevated oldpeak
cleaned_df <- cardio_df %>%
  filter(age < 50, serumcholestrol > 0,serumcholestrol < 200) %>%  # Normal if serumcholestrol < 200 mg/dL
  mutate(
    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),  # Elevated blood pressure: restingBP between 120 and 129
    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")  # Elevated ST depression: oldpeak > 1
  ) %>% 
  select(age, chestpain, restingBP, serumcholestrol, fastingbloodsugar, maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)   # Select variables of interest
    


# Summarize the counts by blood pressure status, ST depression status, and CVD risk
summary_table <- cleaned_df %>%
  group_by(elevated_BP, elevated_oldpeak, target) %>%
  summarise(count = n(), .groups = "drop")

print(summary_table)

# Visualize CVD risk by blood pressure and ST depression status
ggplot(cleaned_df, aes(x = elevated_BP, fill = target)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ elevated_oldpeak) +
  labs(title = "CVD Risk by Blood Pressure and ST Depression Status\n(Individuals Age < 50 with Normal Cholesterol)",
       x = "Blood Pressure Status",
       y = "Count",
       fill = "Heart Disease") +
  theme_minimal()

```
